experDesign: stratifyuing samples into batches with minimal bias

### 2021-07-12 paper.Rmd
Add acknowledgments
Some changes from reviewers

---
title: 'experDesign: helping performing experiments on batches'
output:
  md_document:
    preserve_yaml: yes
  html_document:
    df_print: paged
  word_document:
    keep_md: yes
date: "23 April 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

@@

The design of an experiment is critical to its success.
Nonetheless, even with a correct design, the process up to the moment of measurement is critical.
At any one of the several steps required from collection to measurement various errors and problems could affect the experimental results.
Failure to take such variability into account can render an experiment inconclusive.
*experDesign* provides tools to minimize the risk of inconclusive results by specifically assigning samples to particular batches.

@@

To properly design an experiment, the source of the variation between samples must be identified.
Typically, one can control the environment in which the study or experiment is being conducted.
Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
There are three techniques used to reduce unwanted variation: blocking, randomization and replication [@klaus2015].

**Blocking** group samples that are equal, according to one or more variables, allows one to estimate the variation in the measurements of the factors from these variables.
**Randomization** is a method to minimize the variation in the measurements by mixing the potential confounding variables.
**Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
In some settings (e.g., clinical, agriculture, etc.), several of these techniques can be applied together to ensure the robustness of the study.

Between the designing of an experiment and the measurement of the samples, many things can happen.
If some samples are lost, are contaminated, or do not pass quality control for measurement purposes, the unwanted variation of the experiment will increase.
Even if this doesn't happen, experiments will occasionally need to be carried out in batches due to technical reasons; for example, the machine cannot measure more than one samples at a time, or because of other practical considerations; for instance, it may not be possible to obtain additional measurements in the field during the allotted time.

There are several techniques to identify and assess batch effects when analyzing an already measured experiment [@leek2010].
Experiments that run over long periods of time or that are run across different laboratories are highly susceptible to batch effects.
This is also true of even smaller single-laboratory studies, if they span several days or include personnel changes.
However, if the source of variations are not taken into account before measuring is carried out, a batch effect can be introduced that later is almost impossible to remove.
Thus, it would be better to avoid such batch effects before carrying out an experiment.
Taking into account the process from the design phase to the final measurement can help prevent batch effects.

Experiments should be designed to distribute batches and other potential sources of experimental variation across groups.
To prevent the occurrence of batch effects after the initial design of the experiment, there are two options: randomization and replication.
These techniques can not only prevent batch effects, but can also facilitate comparisons between or within the desired variable(s).
Both can be utilized, but in any case one must take blocking into account from the beginning or the problem might be further exacerbated.

Randomization, if done correctly, can helps reducing variations across groups, while replication can help one be more confident of the experimental results.
For instance, if one designs an experiment with cases and controls and the latter are measured in one batch and the controls in a different batch, then any difference between them can be entirely confounded by a malfunction of the measuring machine in either one of the batches [@chen2011].
However, by examining how the variables are distributed across each batch sample, proper randomized can be ensured, thus minimizing batch effects.
This is known as randomized block experimental design or stratified random sampling experimental design.

Replications consist of increasing the number of measurements with similar attributes.
Usually at least three samples are included for each condition of interest in order to estimate the variations of these attributes.
When there is one extraction and then a sample is measured multiple times, this is referred to as a technical replicate.
Technical replicates help estimate the variation of the measurement method, and thus the possible batch effect.

Replications consist on increasing the number of measurements with similar attributes.
Usually at least three samples are included for each condition of interest to be able to estimate the variation of the attributes.
If there is one extraction and then a sample is measured multiple times it is called a technical replicate.
Technical replicates help estimate the variation of the measurement method and thus of the possible batch effect.

@@

There are some tools to prevent batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
They carry some caveats limiting their application in some cases, and to our knowledge no comparisons of these various methods has been conducted.
Here we briefly describe the currently available packages:

@@

-   Recently, *Omixer*, a new package, has recently been made available at [Bioconductor](https://bioconductor.org/packages/Omixer/) [@sinke2021].
    It tests whether the random assignments are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise , it utilizes the Pearson's chi-squared test.

-   Finally there is the package *experDesign*, at [CRAN](https://cran.r-project.org/package=experDesign), which provides groups with characteristics similar to the entire sample by comparing the random samples with the whole dataset across several statistics.

@@

The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
It is similar to the *anticluster* method used for maximum variance.
If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position in a manner similar to *Omixer*.

In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
When the design is complete, the replicates are referred to as technical replicates [@blainey2014].Technical replicates are samples that are measured multiple times, which reduces the uncertainty of the measurement.
If the technical replicates are distributed across several batches, batch effects can be measured and thus minimized.
To select the technical replicates needed and in order to choose from which samples they are calculated, the function `extreme_cases` is provided.
For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.

@@

>>>>

---
title: 'experDesign: stratifyuing samples into batches with minimal bias'
output:
  html_document:
    df_print: paged
  md_document:
    preserve_yaml: yes
  word_document:
    keep_md: yes
date: "23 April 2021"
affiliations:
- name: Centro de Investigación Biomédica en Red, Enfermedades Hepáticas y Digestivas
  index: 1
- name: Institut d'Investigacions Biomèdiques August Pi i Sunyer, IDIBAPS
  index: 2
authors:
- name: Lluís Revilla Sancho
  orcid: 0000-0001-9747-2570
  affiliation: 1, 2
- name: Juan-José Lozano
  orcid: 0000-0001-9747-2570
  affiliation: '1'
- name: Azucena Salas
  orcid: 0000-0001-9747-2570
  affiliation: '2'
tags:
- R
- batch effect
- experiment design
bibliography: paper.bib
---

```{r setup}
knitr::opts_chunk$set(collapse = TRUE)
```

@@

The design of an experiment is critical to its success.
Nonetheless, even with a correct design, the process up to the moment of measurement is critical.
At any one of the several steps required from collection to measurement various errors and problems could affect the experimental results.
Failure to take such variability into account can render an experiment inconclusive.
*experDesign* provides tools to minimize the risk of inconclusive results by assigning samples to batches to minimize potential batch effects.

@@

To properly design an experiment, the source of the variation between samples must be identified.
Typically, one can control the environment in which the study or experiment is being conducted.
Sometimes, however, this is not possible, and then one needs to apply certain techniques to control for such variations.
There are three techniques used to decrease the uncertainty about the unwanted variation: blocking, randomization and replication [@klaus2015].

**Blocking** groups samples that are equal according to one or more variables allowing one to estimate the differences between each batch by comparing measurements within blocks.
**Randomization** by mixing the potential confounding variables minimizes the variation in the measurements .
**Replication** increases the number of samples used in an experiment to better estimate the variation of the experiment.
In some settings these techniques can be applied together to ensure the robustness of the study.

Between the designing of an experiment and the measurement of the samples, some samples might be lost, be contaminated, or degradation below quality control threshold can happen.
Even if this does not happen, experiments will occasionally need to be carried out in batches.
Due to technical, for example, the machine cannot measure more than one sample at a time; and practical reasons, for instance, it may not be possible to obtain additional measurements in the field during the allotted time.

This divergence from the original design might cause batch effects perturbing the analysis.
There are several techniques to identify and assess batch effects when analyzing an already measured experiment [@leek2010].
However, if the source of variations is not taken into account before measurement is completed, a batch effect can be introduced that confounds the analysis.
Thus, it would be better to avoid such batch effects before executing an experiment.

To prevent the batch effect to confound your analysis after the initial design of the experiment, there are two options: randomization and replication.
Randomization, can help reduce variations across groups, while replication helps estimate the variation of the measurements or samples, thus increasing the precision of the estimates of the true value on the analysis.
Replications consist of increasing the number of measurements with similar attributes.
When there is one extraction and then a sample is measured multiple times, this is referred to as a technical replicate.
Technical replicates help estimate the variation of the measurement method, and thus the possible batch effect.

Randomization and replicates can be utilized to prevent batch effects to confound the analysis.
However, by examining how the variables are distributed across each batch sample, proper randomization can be ensured, thus minimizing batch effects.
This is known as randomized block experimental design or stratified random sampling experimental design.

@@

There are some tools to prevent batch effects on the R language in multiple fields and areas, and particularly for biological research [@rcoreteam2014].
Here we briefly describe the currently available packages:

@@

-   Recently, *Omixer*, a new package, has recently been made available at [Bioconductor](https://bioconductor.org/packages/Omixer/) [@sinke2021].
    It tests whether the random assignments are homogeneous by transforming all variables to numeric values and using the Kendall's correlation when there are more than 5 samples; otherwise, it utilizes the Pearson's chi-squared test.

# Statement of need

Current solutions to stratify samples to reduce and control batch effect do not work for all cases and usual needs of researchers.
They are either specialized to specific type of data or omit some conditions that are often meet or only work on a specific subset of conditions.
The new package ***experDesign*** works on all kind of data types and doesn't require a spatial distribution making it suitable for all kind of experiments.
This package is intended for people looking for a quick easy to interpret solution providing reasonable suggestions about how to distribute the samples on their analysis.

For completeness a description and comparison of the usage of the different software currently available on CRAN and Bioconductor is presented below.
First some real data from a survey:

```{r example-data}
data(survey, package = "MASS")
survey$ID <- seq_len(nrow(survey))
VoI <- c("Sex", "Smoke", "Age")
n_batch <- 3
size_subset <- 96
iterations <- 1000
```

This data set has three variables of interest; Sex, Smoke and Age are a mix of categorical and numeric variables.

## OSAT

OSAT provides some template for some setups and works for numeric and categorical variables:

```{r osat}
library("ggplot2")
library("OSAT")
library("patchwork")
gs <- setup.sample(survey, optimal = VoI)
gc <- setup.container(plate = IlluminaBeadChip96Plate, n = n_batch, batch = 'plates')
gSetup <- create.optimized.setup(sample = gs, container = gc, nSim = iterations)
osat_report <- get.experiment.setup(gSetup)
nrow(osat_report)
osat_age <- ggplot(osat_report) +
  geom_histogram(aes(Age, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
osat_sex <- ggplot(osat_report) +
  geom_bar(aes(Sex, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
osat_smoke <- ggplot(osat_report) +
  geom_bar(aes(Smoke, fill = as.character(plates))) +
  facet_wrap(~plates) +
  labs(fill = "Batch")
(osat_smoke/(osat_age + osat_sex + plot_layout(guides = 'collect'))) +
  plot_annotation(title = "OSAT")
```

OSAT returns one row less that the input provided because it deletes one row because it has an NA value on the Sex variable.

## anticlust

anticlust doesn't handle all the variables, it only accepts numeric variables:

```{r anticlust, error=TRUE}
library("anticlust")
anticlust_index <- anticlustering(
  survey[, VoI],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
anticlust_index <- anticlustering(
  survey[, "Age"],
  K = n_batch,
  objective = "variance",
  method = "exchange",
  repetitions = iterations
)
report_anticlust <- data.frame(Age = survey[, "Age"], batch = as.character(anticlust_index))
library("ggplot2")
ggplot(report_anticlust) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch", title = "anticlust")
```

## Omixer

There is a bug on Omixer that prevents it to work unless specific conditions are meet that prevents it comparing with the other tools using the same settings:

```{r Omixer, error=TRUE}
library("Omixer")
data(survey, package = "MASS")
survey$ID <- seq_len(nrow(survey))
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, 
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, nrow(survey)),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(1, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)
Omixer_index <- omixerRand(survey, sampleId = "ID", 
                           block = "block", iterNum = 10, wells = size_subset, mask = rep(0, 288),
                           plateNum = n_batch, randVars = VoI, techVars = NULL)

```

## experDesign

***experDesign***, the package described on this paper, first evaluates the distribution of samples in batches and then on a spatial grid.

```{r experDesign}
library("experDesign")
experDesign_index <- design(survey[, VoI], size_subset = size_subset,
                  iterations = iterations)
experDesign_index_spatial <- spatial(index = experDesign_index, pheno = survey[, VoI], 
                                     iterations = iterations, rows = LETTERS[1:8], columns = 1:12)
experDesign_report <- inspect(experDesign_index, survey[, VoI])
nrow(experDesign_report)
experDesign_age <- ggplot(experDesign_report) + 
  geom_histogram(aes(Age, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_sex <- ggplot(experDesign_report) +
  geom_bar(aes(Sex, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
experDesign_smoke <- ggplot(experDesign_report) +
  geom_bar(aes(Smoke, fill = batch)) +
  facet_wrap(~batch) +
  labs(fill = "Batch")
(experDesign_smoke/(experDesign_age + experDesign_sex + plot_layout(guides = 'collect'))) + plot_annotation(title = "experDesign")
```

@@

The package ***experDesign*** provides the functional `design` to distribute the samples into multiple batches such that each variable is homogeneous within each batch.
Each batch is set to have some centrality and dispersion statistics to match as closely as possible the original input design data.
It does by comparing the mean, the standard deviation, the median absolute deviation, variables with no value number, the entropy and the independence of the categorical variables.
The distribution of samples that more closely match the original distribution is returned to the user.

This approach is similar to the *anticlust* method used for maximum variance, where the objective function is the sum of the squared errors between cluster centers and individual data points.
But instead of comparing to the clusters center it is compared to the whole data.
If the experiment is carried out with a specific spatial distribution, the `spatial` function also distributes the samples homogeneously by position.

In addition to distributing the samples into batches, ***experDesign*** provides tools to add technical replicates.
When the design is complete, the replicates are referred to as technical replicates [@blainey2014].
To select the technical replicates needed and in order to choose from which samples they are calculated, the function `extreme_cases` is provided.
For easier usage, the `replicates` function designs an experiment with the number of replicates per batch desired.

```{r controls}
extreme_cases(survey[, VoI], size = 5)
replicates_index <- replicates(survey[, VoI], size_subset = size_subset, controls = 5, iterations = iterations)
spatial_replicate_index <- spatial(replicates_index, survey[, VoI])
inspect(spatial_replicate_index, survey[, VoI])
```

@@

```{r show-batches}
optimum_batches(size_data = 250, size_subset = 96)
sizes_batches(size_data = 250, size_subset = 96, batches = 3)
```

# Acknowledgments

We are grateful to Joe Moore for English-language assistance.
